A real-time particle filter for modeling travel times of transit vehicles through a GTFS-based road network

### 2018-10-30 anzjs2018/anzjs2018.tex
Write some words about moels
more writing
Writing model information
models
network model
Implementation info
intro very drafty
Double spacing
Figures for network and pf
edits

\documentclass[times, doublespace]{anzsauth}
\usepackage{moreverb}
\usepackage{url}
\usepackage{grffile}
\usepackage{lineno}
% \usepackage{lmodern}
\usepackage{lipsum}
\usepackage[UKenglish]{isodate}

@@

@@

\begin{document}
\cleanlookdateon
\runningheads{Short title of paper}{TOM~ELLIOTT AND THOMAS~LUMLEY}
\title{Long name of the paper}
\author{Tom Elliott\corrauth~and~Thomas~Lumley}
\affiliation{University of Auckland}
\address{
    Department of Statistics, University of Auckland, 
    Auckland 1142, New Zealand\\
    Email: \texttt{tom.elliott@auckland.ac.nz}
}

\begin{abstract}
Predicting transit vehicle arrival time is a somewhat complex procress.
Road networks are dynamic, and can change from free-flowing to highly congested
very quickly.
Any reliable prediction framework needs to be able to respond to these changes,
but also to future trends, such as typical changes to traffic flow
at peak times, based on hisorical data.
Of course, the major constraint on any prediction framework is going to be
computational efficiency.
Due to the realtime nature of the problem, predictions need to be 
available as soon as possible after observing the vehicles' positions,
which are updated approximately every 30~seconds.
In this paper, we describe our C++ implementation of a particle filter to estimate
vehicle state in combination with a network state model,
allowing us to model transit vehicle flow through the network.
While our application here is being testing in Auckland, New Zealand,
we hope that our methods are general enough to be applied to any
transit network that uses GTFS.
Preliminary results show that, with suitable hardware,
the framework is fast enough while retaining the complexity necessary
to incorporate as much information as necessary to predict arrival times.
\end{abstract}

\keywords{particle filter; Kalman filter; transit modeling;
          transit networks; statistical computing; applications; gtfs}

\maketitle
\input{sections/01_introduction}
\input{sections/02_gtfs}
\input{sections/03_models}
\input{sections/04_implement}
\input{sections/05_results}
\input{sections/06_discuss}
\bibliographystyle{anzsj}
\bibliography{../reflist.bib}

>>>>

\documentclass[times, doublespace]{anzsauth}
\usepackage{moreverb}
\usepackage{url}
\usepackage{grffile}
\usepackage{lineno}
% \usepackage{lmodern}
\usepackage{lipsum}
\usepackage[UKenglish]{isodate}
\usepackage{float}
\usepackage{subcaption}
\usepackage{algorithm}
\usepackage{algpseudocode}
% \newfloat{lstfloat}{htbp}{lop}
% \floatname{lstfloat}{Listing}
% \def\lstfloatautorefname{Listing}

% \doublespace

@@

\newcommand{\bv}{\boldsymbol{v}}
\newcommand{\bz}{\boldsymbol{z}}
\newcommand{\bw}{\boldsymbol{w}}
\newcommand{\bR}{\boldsymbol{R}}
\newcommand{\bT}{\boldsymbol{T}}
\newcommand{\br}{\boldsymbol{r}}
\newcommand{\btheta}{\boldsymbol{\theta}}

\newcommand{\rt}{real-time\ }

@@

\begin{document}
\cleanlookdateon
\runningheads{Modeling transit vehicles in real-time}{TOM~ELLIOTT AND THOMAS~LUMLEY}
\title{A real-time particle filter for modeling travel times of transit vehicles through a GTFS-based road network}
\author{Tom Elliott\corrauth~and~Thomas~Lumley}
\affiliation{University of Auckland}
\address{
    Department of Statistics, University of Auckland, 
    Auckland 1142, New Zealand\\
    Email: \texttt{tom.elliott@auckland.ac.nz}
}

\begin{abstract}
Predicting the arrival time of a transit vehicle involves not only
knowledge of its current position and schedule adherance,
but also information about traffic conditions along the remainder of its route.
Road networks are dynamic, 
and can change from free-flowing to highly congested very quickly,
which can effect the arrival time of transit vehicles, 
buses in particular,
% This in turn leads to commuter frustration when they have to wait longer than
% initially anticipated,
so any reliable prediction framework needs to account for \rt traffic conditions.
In this paper,
we propose a particle filter that can be used to estimate travel times
along roads within a transit network based solely on GTFS data.
Due to the \rt nature of the problem and computational demands of such an approach,
we demonstrate the faesibility of our program, 
written in the R programming language and making extensive use of C++.
The application provides the basis from which future work will be based,
most notably for combining the \rt travel times with historical data and a 
formal road network model to make reliable arrival time predictions.
% Of course, the major constraint on any prediction framework is going to be
% computational efficiency.
% Due to the realtime nature of the problem, predictions need to be 
% available as soon as possible after observing the vehicles' positions,
% which are updated approximately every 30~seconds.
% In this paper, we describe our C++ implementation of a particle filter to estimate
% vehicle state in combination with a network state model,
% allowing us to model transit vehicle flow through the network.
Development and testing has taken place here in Auckland, New Zealand,
although we expect our methods are general enough to be applied to any
transit network using GTFS.
\end{abstract}

\keywords{particle filter; transit modeling;
          transit networks; applications; gtfs}

\maketitle
\input{sections/01_introduction}
\input{sections/02_gtfs}
\input{sections/03_models}
\input{sections/04_implementation}
% \input{sections/05_results}
\input{sections/06_discuss}
\bibliographystyle{anzsj}
\bibliography{../reflist.bib}

### 2018-10-30 anzjs2018/sections/01_introduction.tex
edits

In public transport, \rt information (RTI), for example estimated times of arrival (ETAs),
keeps transport users informed about their journey,
and in some places make decisions on the best route to take.
Unfortunately, here in Auckland, RTI is highly unreliable,
and succumbs to several problems that public transport users
are all too familiar with.
One of these is inaccurate ETAs,
which tend to either increase or decrease over time as the bus gets progressively later or earlier
due to traffic congestion, or  because the schedule is poorly calibrated.
A second is that buses are shown as on time but in fact are not running,
and therefore have not registered on the \rt system.
In this case, the scheduled times are used for ETAs and, 
when the bus doesn't show up, 
the bus disappeared from the \rt board and we get what passengers have described as ``Ghost Buses''.

Arrival time prediction is only as accurate as the underlying model,
and while a lot of work has gone into developing public transport vehicle models
\citep{Cathey_2003,Jeong_2005,Yu_2011,Hans_2015},
in many public transport systems no formal model is used.
Instead, ETAs are solely based on the scheduled arrival time
adjusted by the vehicle's delay at the most recent stop, if available.
This assumes that the schedule is valid,
and that there is no unusual congestion along the route,
neither of which are valid assumptions,
particularly in our test area of Auckland, New Zealand,
where infrastructure for buses (such as priority lanes) is limited.
This behaviour is exaggerated when there is a large distance between stops,
as the ETA can change drastically.

To avoid these problems and improve the reliability of ETAs,
we first needed to develop a model to infer position and speed
of transit vehicles from \rt GPS data,
isntead of relying on delays reported only at stops.
Several modeling approaches were explored, 
such as the Kalman filter (used by \cite{Dailey_2001} and \cite{Cathey_2003}, ...),
machine learning models \citep{Yu_2006,Chang_2010},
and a particle filter approach demonstrated
by \cite{Hans_2015}.
We chose to use the last of these, the particle filter,
as they are very flexible and aren't limited by assumptions
such as unimodality;
instead, they are constrained by computational complexity.

The \emph{reliablility} of arrival time estimates
alo needs to be considered.
Ideally, congestion along intermediate roads should be incorporated into the predictive model,
and one way of doing this is to use the travel time of preceeding buses along the same roads.
Several such approaches have been proposed.
\cite{Yu_2010} used the travel time of vehilces on the same \emph{routes},
which they showed to improve ETAs.
However, this only works on routes with a high frequency.
\citep{Yu_2011} extended this by looking at roads where several major rouets converged,
again showing improvements but limited in scope.
In general, however, no comprehensive network modeling approach has been proposed that used
solely GPS position data to estimate arrival times
while accounting for congestion.

In this paper, we describe a \rt vehicle and network state
model which uses the vehicles traveling throughout the transit road network
to update congestion information in \rt.
The vehicle model is implemented using a particle filter,
so it is also necessary to demonstrate the faesibility of this as a valid real-time option.
From the particle filter estimates of vehicle trajectories,  
we estimate the travel time along roads update the road network's state.
Future work will involve using this travel time information 
in combination with historical data to develop an arrival time prediction model.

>>>>

In public transport, \rt information (RTI), for example estimated times of arrival (ETAs),
keeps transport users informed about their journey,
and in some places make decisions on the best route to take.
Unfortunately, here in Auckland, RTI is highly unreliable,
and succumbs to several problems that public transport users
are all too familiar with.
One of these is inaccurate ETAs,
which tend to either increase or decrease over time as the bus gets progressively later or earlier
due to traffic congestion, or  because the schedule is poorly calibrated.
Another is that buses are shown as \emph{on-time} if they are not present
in the \rt system,
no matter whether this is due to technical problems,
or that the bus has failed to show up;
in this case, schedule times are used to determine ETAs and, 
if the bus doesn't show up on time,
the service is removed from the \rt board 
and we get what passengers have described as ``Ghost Buses''.

Arrival time prediction is only as accurate as the underlying model,
and while a lot of work has gone into developing public transport vehicle models
\citep{Cathey_2003,Jeong_2005,Yu_2011,Hans_2015},
in many public transport systems no formal model is used.
Instead, ETAs are solely based on the scheduled arrival time
adjusted by the vehicle's delay at the most recent stop, if available.
This assumes that the schedule is valid,
and that there is no unusual congestion along the route,
neither of which are valid assumptions,
particularly in our test area of Auckland, New Zealand,
where infrastructure for buses (such as priority lanes) is limited.
This behaviour is exaggerated when there is a large distance between stops,
as the ETA is only updated at stops so there can be a long time between updates.

To avoid these problems and improve the reliability of ETAs,
we required two things:
first, a robust vehicle model to estimate the position and speed
of transit vehicles from \rt GPS data;
and secondly, a way to incorporate road congestion information 
into the arrival time predictions.
Several vehicle modeling approaches were explored, 
such as the Kalman filter (used by \cite{Dailey_2001} and \cite{Cathey_2003}, ...),
machine learning models \citep{Yu_2006,Chang_2010},
and a particle filter approach demonstrated
by \cite{Hans_2015}.
The last of these, the particle filter, was chosen for several reasons,
discussed in Section~\ref{sec:pf}.

In order to improve the reliability of ETAs,
congestion along intermediate roads should be incorporated into the predictive model.
One solution is to use the travel time of preceeding buses along the same roads
to predict travel times for the current bus,
such as proposed by \cite{Yu_2010},
who were able to improve prediction accuracy
by using the travel time of vehicles on the same \emph{route}.
However, this only works well for high-frequency routes, so
\cite{Yu_2011} modeled travel times along roads where several major routes converged,
which showed further improvement however is limited in scope.
In general, however, no comprehensive network modeling approach has been proposed us
solely GPS position data to estimate arrival times
while accounting for congestion.

In this paper, we describe a \rt vehicle and network state
model which uses the vehicles traveling throughout the transit road network
to update congestion information in \rt.
As the vehicle model is implemented using a particle filter,
we demonstrate its faesibility as a valid real-time option.
From the particle filter estimates of vehicle trajectories,  
The particle filter provides estimates of vehicle travel times along roads,
which can be used to update the road network's state.
Future work involves using this travel time information 
in combination with historical data to develop an arrival time prediction model.

### 2018-10-30 anzjs2018/sections/02_gtfs.tex
edits

GTFS (general transit feed specification)
is an API (application programming interface) specification for transit data
detailing how it should be organised,
making access easier for application developers.
Developed and maintained by Google \citep{GoogleDevelopers_2006},
who make use of it in their Google Maps Transit Directions,
it is used by over 900~transit providers around the world,
including here in Auckland, New Zealand
(source \url{http://transitfeeds.com}).
An advantage of this standardised format is that,
provided an application depends solely on GTFS data,
after developing it locally in Auckland it can be deployed to any other GTFS-based
public transport system with minimal modification.

@@

As mentioned in Section~\ref{sec:intro},
there are some major issues with the current ETA generation method.
These are almost directly attributed to GTFS-realtime trip updates,
which are currently the sole source of data for ETAs (in Auckland).
Trip updates are reported whenever a vehicle arrives or departs a stop,
and includes most importantly the delay between the scheduled and actual arrival times.
This delay is then propagated on to all future stops to adjust the ETA.
This assumes that the schedule is well calibrated and the time between scheduled arrivals
is representative of the real-world travel time between stops. 
Even more problematic is that, in the absense of trip updates,
the delay defaults to zero,
and unless the trip is manually cancelled,
appears on-time to passengers waiting at the stop.
If the bus is late or cancelled,
the trip disappears from the \rt board and passengers
are unable to distingush the two scenarios.

@@

Arguably the most important predictor of arrival time is
the cumulative travel time along intermediate roads.
In most applications, however, this vital information is unavailable,
at least directly,
so here we attempt to construct a transit road network of intersections
and the roads between them, 
and map each route to a sequence of road segments.

The simplest way to construct such a network is to use the stop sequence:
routes with a common sub-sequence of stops must be traveling the same route between them,
so by defining a road as the path between two consecutive stops 
we can obtain travel time estimates along that road segment.
There are several places where this doesn't quite work, for example express routes 
and multi-stop locations where there are high numbers of routes,
resulting in some overlap between road segments,
it is a viable enough simplication for the purposes of this paper.
Figure~\ref{fig:network_creation} demonstrates how several overlapping routes 
are merged to form a road network.
In section~\ref{sec:kf} we present a model for estimating vehicle travel time
along a road segment.
Future work will look at using physical intersections as nodes in the network to improve it.

@@

GTFS \rt allows developers to query the current positions of vehicles
in the transit network.
The data consists of the time $t_k$ that the observation was made,
the GPS position of the vehicle, $\bY_k$, 
and some other information about the trip being serviced.
Vehicle positions are updated with a frequency of anywhere between 10~seconds and several minutes,
so there is often a lot of uncertainty about the trajectory
between two observations, particularly when there is one or more bus stop
or intersection between them.
It is also possible for a bus to remain stationary,
so the possible vehicle positions rapidly increases with the time between observations.

Another complication with the Auckland Transport realtime feed is that
the buses are programmed to report their location when arriving at
bus stops and some major intersections.
Often these updates are preemptive 
(i.e., the bus is almost there, but not quite),
and subsequent observations place the bus \emph{behind} the stop or intersection
(e.g., in a queue of traffic at traffic lights).
To handle this, we compute the approximate distance traveled, $\tilde x_k$,
of the vehicle by finding the nearest point on the path to the observation;
if this has decreased, the current state is rejected and the vehicle reverted
to its previous state.

>>>>

GTFS (general transit feed specification)
is an API (application programming interface) specification for transit data
detailing how it should be organised,
making access easier for application developers.
Developed and maintained by Google \citep{GoogleDevelopers_2006},
who use it in Google Maps Transit Directions,
it is used by over 900~transit providers around the world,
including here in Auckland, New Zealand
(source \url{http://transitfeeds.com}).
An advantage of this standardised format is that,
provided an application depends solely on GTFS data,
after developing it locally in Auckland it can be deployed to any other GTFS-based
public transport system with minimal modification.

@@

As mentioned in Section~\ref{sec:intro},
there are some major issues with the current prediction method in Auckland.
These are almost directly attributed to GTFS-realtime trip updates,
which are currently the sole source of data for ETAs.
Trip updates are reported whenever a vehicle arrives or departs a stop,
and includes the delay between the scheduled and actual arrival times,
and is propagated to all future stops to adjust their ETAs.
This assumes that the schedule is well calibrated and the time between scheduled arrivals
is representative of the real-world travel time between stops. 
In many places in Auckland, 
traffic congestion is highly dynamic and changes from day to day,
so even a well-calibrated schedule is an unreliable baseline.

@@

The primary predictor of arrival time is 
the travel time along roads between where the bus is \emph{now},
and the stop at which the passenger is waiting.
In most applications, however, this vital information is unavailable,
at least directly,
so here we construct a transit road network consisting of intersections
and the roads between them.
By generalising to physical roads,
we remove any route-dependency,
which increases the quantity of data available for each road.
Then, each route can instead be represented as a sequence of road segments,
each with a travel time state.

To construct the network,
we use stops as nodes, so that
routes with a common sub-sequence of stops will share the roads between them;
roads are then defined as the links between stops,
as demonstrated in Figure~\ref{fig:network_creation}.
In section~\ref{sec:kf} we present a model for estimating vehicle travel time
along a road segment.
There are several places where this doesn't quite work, 
for example express routes which do not service certain stops,
and multi-stop locations,
which results in overlapping road segments.
Future work will look at using intersection points of the route shapes
as nodes in the network to improve the coverage and reduce overlap.

@@

GTFS-realtime provides the current positions of vehicles in a transit network.
The data consists of the time $t_k$ that the observation was made,
the GPS position of the vehicle, $\bY_k$, 
and information about the trip being serviced.
In Auckland, vehicle positions are updated with a frequency of anywhere between 10~seconds and several minutes,
so there is often a lot of uncertainty about a vehicle's intermediate trajectory,
particularly when there are one or more bus stops along the way.
It is also possible for a bus to remain stationary,
for example due to heavy congestion,
so the number of possible trajectories rapidly increases with 
the time between observations.

One important consideration regarding the Auckland Transport's \rt implementation is that
buses are programmed to report their location when arriving at or departing from
bus stops, as well as some major intersections.
To further complicate this,
these positions can be preemtive,
(i.e., the bus is almost there, but not quite),
so consecutive observations can show a bus traveling backwards.
To handle this, we compute the approximate distance traveled, $\tilde x_k$,
of the vehicle by finding the nearest point on the path to the observation;
if this has decreased, the current state is rejected and the vehicle reverted
to its previous state become continuing.
This way, we avoid degeneration of the particle fitler (Section~\ref{sec:model_perf})
at the cost of losing one observation that is most likely invalid anyway.
